A full approximation scheme multilevel method \\ for nonlinear variational inequalities

### 2023-07-20 paper/fascd.tex
Some edits to section 4

In the material which follows, use of higher-order Lagrange elements, i.e.~$P_k$ or $Q_k$ with $k\ge 2$, would imply a ``variational crime'' \cite[Chapter 10]{BrennerScott2007} special to inequality constraints, which we avoid.  Namely, our FE construction of constraint sets (Section \ref{sec:cdmultilevel}) will require \emph{nodal monotonicity},
\begin{equation}
\bw \ge \bz \quad \implies \quad w \ge z \label{eq:nodalmonotonicity}
\end{equation}
for all $w,z \in \mathcal{V}^j$, where $\bw,\bz \in \RR^{m_j}$ denote vectors of nodal values.  However, for $k\ge 2$ there are basis functions in $P_k$ and $Q_k$, i.e.~functions with nodal values one or zero, which take on negative values \cite[Figure 1.7]{Elmanetal2014}.  Evidently \eqref{eq:nodalmonotonicity} is equivalent to the non-existence of such indefinite-sign basis functions.
\pef{I'm not sure this constitutes a variational crime? We enforce the RHS because it's tractable, and this is only equivalent to what we actually want (the LHS) for P1 and Q1. I don't see where we're departing from the variational framework, however.}

To allow box constraints on all meshes in the hierarchy, let $\tilde{\mathcal{V}}^j$ denote the set of functions on the vertices of $\mathcal{T}^j$ with values in the extended real line $\tilde{\RR} = [-\infty,+\infty]$.  Suppose that the finest-level \emph{lower} and \emph{upper obstacles} $\underline{\gamma}^J, \overline{\gamma}^J \in \tilde{\mathcal{V}}^J$ satisfy
\begin{equation}
\underline{\gamma}^J \le \overline{\gamma}^J, \qquad \underline{\gamma}^J < +\infty, \qquad -\infty < \overline{\gamma}^J. \label{eq:fe:boxconstraintrequirements}
\end{equation}
\pef{Shouldn't there be some relation between the continuous bounds and the discrete ones?}
In analogy to the continuous data, also assume that $\underline{\gamma}^J \le g_D^J \le \overline{\gamma}^J$ on those vertices of $\mathcal{T}^J$ which lie in $\partial_D \Omega$.  Define
\begin{equation}
\mathcal{K}^J = \left\{w\,:\,\underline{\gamma}^J \le w \le \overline{\gamma}^J \text{ on vertices of } \mathcal{T}^J, \, \text{ and } \, w|_{\partial_D\Omega} = g_D^J|_{\partial_D\Omega}\right\} \subset \mathcal{V}^J, \label{eq:fe:fineconstraintset}
\end{equation}
the finest-level \emph{constraint set} of admissible FE functions. This is a nonempty, closed, and convex subset of $\mathcal{V}^J$.

@@

While \eqref{eq:fe:vi} approximates \eqref{eq:boxdirichletvi}, we warn the reader that it may not technically be conforming for that VI.  That is, even if quadrature and other implementation details for $f^J$ are ignored, and despite the vector space inclusion $\mathcal{V}^J \subset \mathcal{V}$, we do not require or generally expect $\mathcal{K}^J \subset \mathcal{K}$.  Regardless of the smoothness of the continuum obstacles $\underline{\gamma}, \overline{\gamma}$, admissibility with respect to their FE interpolants $\underline{\gamma}^J, \overline{\gamma}^J$ will generally not imply continuum admissibility.\footnote{This caveat is especially relevant in geophysical situations where only nodal values of the obstacle are known.  Nodal elevation data generally misses the actual topographic detail in-between the vertices; compare \cite{Bueler2016}.}

@@

The monotone injections $\maxR,\minR$ \cite{GraeserKornhuber2009}, which will only be applied to nodal functions $\tilde{\mathcal{V}}^j$, are nonlinear maps which maximize and minimize, respectively, nodal values over the open supports of coarser-mesh basis functions (i.e.~over the \emph{stars} of the coarse vertices). These operators are constructed to enforce the property
\begin{equation}
\minR z \le z \le \maxR z,  \label{eq:fe:monotonerestrictionprops}
\end{equation}
which does not hold for the standard injection operator $\iR$.
To make the construction precise, let $x_q^j$ denote the vertices of mesh $\mathcal{T}^j$, with corresponding basis functions $\psi_q^j(x)$.  For a $\tilde\RR$-valued nodal function $z\in\tilde{\mathcal{V}}^j$ we define functions in $\tilde{\mathcal{V}}^{j-1}$ (Figure \ref{fig:Rplusminus}):
\begin{align}
(\maxR z)(x_p^{j-1}) &= \max \{z(x_q^j) \,:\, \psi_p^{j-1}(x_q^j) > 0\}, \label{eq:fe:monotonerestrictions} \\
(\minR z)(x_p^{j-1}) &= \min \{z(x_q^j) \,:\, \psi_p^{j-1}(x_q^j) > 0\}. \notag
\end{align}
For standard refinement a coarse-mesh node $x_p^{j-1}$ is in the set of fine vertices $x_q^j$ such that $\psi_p^{j-1}(x_q^j)>0$, so \eqref{eq:fe:monotonerestrictionprops} follows.
On the other hand, $z\ge 0$ implies $\minR z \ge 0$ and $z \le 0$ implies $\maxR z \le 0$.  These ordering properties are used in the CD construction of the next two Sections.

>>>>

As is common in the discretization of variational inequalities, we employ lowest-order continuous Lagrange elements $P_1$ or $Q_1$ on simplices and tensor-product cells respectively. This choice is convenient because this choice enjoys \emph{nodal monotonicity}:
\begin{equation}
\bw \ge \bz \quad \implies \quad w \ge z \label{eq:nodalmonotonicity}
\end{equation}
for all $w,z \in \mathcal{V}^j$, where $\bw,\bz \in \RR^{m_j}$ denote vectors of nodal values. In other words, to enforce an inequality constraint, with $P_1$ or $Q_1$ we only need enforce it pointwise on the degrees of freedom. For $k\ge 2$ there are functions in the standard choices of bases for $P_k$ and $Q_k$ which take on negative values \cite[Figure 1.7]{Elmanetal2014}, violating \eqref{eq:nodalmonotonicity}.

To allow box constraints on all meshes in the hierarchy, let $\tilde{\mathcal{V}}^j$ denote the set of functions on the vertices of $\mathcal{T}^j$ with values in the extended real line $\tilde{\RR} = [-\infty,+\infty]$.  The finest-level \emph{lower} and \emph{upper obstacles} $\underline{\gamma}^J, \overline{\gamma}^J \in \tilde{\mathcal{V}}^J$ are constructed by suitable interpolation of $\underline{\gamma}$, $\overline{\gamma}$ (e.g.~by applying the Cl\'ement interpolation operator) and satisfy
\begin{equation}
\underline{\gamma}^J \le \overline{\gamma}^J, \qquad \underline{\gamma}^J < +\infty, \qquad -\infty < \overline{\gamma}^J. \label{eq:fe:boxconstraintrequirements}
\end{equation}
Analogously, we construct the discrete boundary data $g_D^J$ so that $\underline{\gamma}^J \le g_D^J \le \overline{\gamma}^J$ on those vertices of $\mathcal{T}^J$ which lie in $\partial_D \Omega$.  Define
\begin{equation}
\mathcal{K}^J = \left\{w\,:\,\underline{\gamma}^J \le w \le \overline{\gamma}^J \text{ on vertices of } \mathcal{T}^J, \, \text{ and } \, w|_{\partial_D\Omega} = g_D^J|_{\partial_D\Omega}\right\} \subset \mathcal{V}^J, \label{eq:fe:fineconstraintset}
\end{equation}
the finest-level \emph{constraint set} of admissible FE functions. This is a nonempty, closed, and convex subset of $\mathcal{V}^J$.

@@

While \eqref{eq:fe:vi} approximates \eqref{eq:boxdirichletvi}, we warn the reader that it is usually not conforming for that VI.  That is, even if quadrature and other implementation details for $f^J$ are ignored, and despite the vector space inclusion $\mathcal{V}^J \subset \mathcal{V}$, we do not require or expect $\mathcal{K}^J \subset \mathcal{K}$.  Regardless of the smoothness of the continuum obstacles $\underline{\gamma}, \overline{\gamma}$, admissibility with respect to their FE interpolants $\underline{\gamma}^J, \overline{\gamma}^J$ will generally not imply continuum admissibility.\footnote{This caveat is especially relevant in geophysical situations where only nodal values of the obstacle are known.  Nodal elevation data generally misses the actual topographic detail in-between the vertices; compare \cite{Bueler2016}.}

@@

The monotone injections $\maxR,\minR$ \cite{GraeserKornhuber2009}, which will only be applied to nodal functions $\tilde{\mathcal{V}}^j$, are nonlinear maps which maximize and minimize, respectively, nodal values over the open supports of coarser-mesh basis functions (i.e.~over the \emph{stars} of the coarse vertices), Figure \ref{fig:Rplusminus}. These operators are constructed to enforce the property
\begin{equation}
\minR z \le z \le \maxR z,  \label{eq:fe:monotonerestrictionprops}
\end{equation}
which does not hold for the standard injection operator $\iR$.
To make the construction precise, let $x_q^j$ denote the vertices of mesh $\mathcal{T}^j$, with corresponding basis functions $\psi_q^j(x)$.  For a $\tilde\RR$-valued nodal function $z\in\tilde{\mathcal{V}}^j$ we define functions in $\tilde{\mathcal{V}}^{j-1}$:
\begin{align}
(\maxR z)(x_p^{j-1}) &= \max \{z(x_q^j) \,:\, \psi_p^{j-1}(x_q^j) > 0\}, \label{eq:fe:monotonerestrictions} \\
(\minR z)(x_p^{j-1}) &= \min \{z(x_q^j) \,:\, \psi_p^{j-1}(x_q^j) > 0\}. \notag
\end{align}
For standard refinement a coarse-mesh node $x_p^{j-1}$ is in the set of fine vertices $x_q^j$ such that $\psi_p^{j-1}(x_q^j)>0$, so \eqref{eq:fe:monotonerestrictionprops} follows.
On the other hand, $z\ge 0$ implies $\minR z \ge 0$ and $z \le 0$ implies $\maxR z \le 0$.  These ordering properties are used in the CD construction of the next two Sections.

### 2023-07-20 paper/tikz/Rplusminus.tex
Some edits to section 4

  }
  \filldraw[seabornblue] (0, 0) circle (1.5pt);
  \node (expl) at (1.5, 0.5) {};
  \filldraw[seabornblue] (expl) circle (1.5pt);
  \node at (2.25, 0.525) {$= \text{max/min of }$};  % I'm sure there's a better way to place this, but tikz mystifies me
  \filldraw[seaborngreen] (2.95, 0.5) circle (1.5pt);
\end{tikzpicture}

>>>>

  }
  \filldraw[seabornblue] (0, 0) circle (1.5pt);
  \node (expl) at (1.5, 0.5) {};
  \filldraw[seabornblue] (expl) circle (1.5pt);
  \node at (2.9, 0.5125) {$= \text{max/min over patch of cells}$};  % I'm sure there's a better way to place this, but tikz mystifies me
  \node at (2.25, 0.2925) {$= \text{max/min of }$};  % I'm sure there's a better way to place this, but tikz mystifies me
  \filldraw[seaborngreen] (2.95, 0.29) circle (1.5pt);
\end{tikzpicture}
